---
title: "Data Visualization Final"
output: html_notebook
---

```{r}
library(dslabs)
library(ggplot2)
library(dplyr)
library(segmented)
library(gridExtra)
diseases <- get("us_contagious_diseases")
murders <- get("murders")
diseases
murders
summary(diseases)
```
Are these reported new cases or all cases?

Replace 0's with NA's
Delete Polio Data
```{r}
diseases[diseases$weeks_reporting == 0, c("weeks_reporting", "count")] <- NA
diseases <- diseases[which(diseases$disease != "Polio"),]
diseases$disease <- factor(diseases$disease)
```

Were weeks that were not reported 0 cases or can we assume that the average reported week applied to each week:
```{r}
Measles <- (diseases[which(diseases$disease == "Measles"),2:ncol(diseases)])
str(Measles)
plot(count ~ year, data=Measles[which(Measles$state == "Alaska"),])
# this is nice because it gets rid of not reported years
# but it doesn't seem to change the overall trend so that's good
Measles$normalized.count <- Measles$count / Measles$weeks_reporting * 52
str(Measles)
plot(normalized.count ~ year, data=Measles[which(Measles$state == "Alaska"),])
```

What about the country?
```{r}
plot(normalized.count ~ year, data=Measles)
# need to normalize by population
Measles$normalized.count.bycapita <- Measles$normalized.count / Measles$population
Measles
plot(normalized.count.bycapita ~ year, data=Measles)
```

Here we can see that even when normalized by population, some areas go hit more than others.
Do more densley populated areas get sick more?

```{r}
Measles[order(Measles$normalized.count.bycapita),]
```

What do we do with states that barely reported?
Do I want to show overall counts for a year or the per capita counts?

Story - the introduction of vaccines

Make a combined dataset will all the counts of a disease per year as first figure.

All diseases
```{r, fig.width=10, fig.height=6}
diseases$disease <- factor(diseases$disease, levels = c("Smallpox", "Measles", "Pertussis", "Hepatitis A", "Rubella", "Mumps"))
diseases$normalized.count <- diseases$count / diseases$weeks_reporting * 52
diseases$percentpop <- diseases$normalized.count/diseases$population
```

```{r, fig.width=10, fig.height=6}
ggplot(na.omit(diseases), aes(x=year, y=normalized.count, color=disease)) + geom_point(alpha=0.2)
ggplot(na.omit(diseases), aes(x=year, y=normalized.count, color=disease)) + geom_point(alpha=0.2) + facet_grid(disease ~ .)
ggplot(na.omit(diseases), aes(x=year, y=normalized.count, color=disease)) + geom_point(alpha=0.2, show.legend = F) + facet_wrap(~ disease, scales="free")
ggplot(na.omit(diseases), aes(x=year, y=normalized.count, color=disease)) + geom_boxplot(aes(group=factor(year)),show.legend = F, size=0.2, outlier.size = 0.1) + facet_wrap(~ disease, scales="free")
```

If I want to look only at the progression of the disease compared to when the vaccine was released, I want to normalize the values by the maximum of each disease.

```{r}
maximums <- aggregate(normalized.count~disease, diseases, FUN=max)
diseases$maximums <- maximums[diseases$disease, "normalized.count"]
diseases$disease.proportion <- diseases$normalized.count/diseases$maximums

ggplot(na.omit(diseases), aes(x=year, y=disease.proportion, color=disease)) + geom_point(alpha=0.1) + geom_smooth(se=FALSE)
```
```{r, fig.width=6, fig.height=4}
diseasesbyyear <- na.omit(diseases) %>%
  group_by(disease,year) %>%
  summarize(totalcount = sum(normalized.count, na.rm = TRUE))

maximums2 <- aggregate(totalcount~disease, diseasesbyyear, FUN=max)
diseasesbyyear$maximums <- maximums2[diseasesbyyear$disease, "totalcount"]

diseasesbyyear$percentofhighest <- diseasesbyyear$totalcount/diseasesbyyear$maximums

ggplot(na.omit(diseasesbyyear), aes(x=year, y=percentofhighest, color=disease)) + geom_point(alpha=0.4, aes(size=totalcount)) + geom_smooth(se=FALSE, size=0.4)
```
```{r}
alldiseases <- diseases %>%
  group_by(year) %>%
  summarize(total = sum(count, na.rm = TRUE))

ggplot(na.omit(alldiseases), aes(x=year, y=total)) + geom_point(alpha=0.4) + geom_smooth(se=FALSE, size=0.2)
```


```{r}
hepA.vaccine <- 1995          #first approved in US
measles.vaccine <- 1963
mumps.vaccine <- 1948         #first licensed, however short term effectiveness, improved in the 1960s
pertussis.vaccine <- 1926
rubella.vaccine <- 1969
smallpox.vaccine <- 1796

vline.data <- data.frame(disease=c("Hepatitis A", "Measles", "Mumps", "Pertussis", "Rubella", "Smallpox"), vl=c(hepA.vaccine, measles.vaccine, mumps.vaccine, pertussis.vaccine, rubella.vaccine, smallpox.vaccine))
ggplot(na.omit(diseases), aes(x=year, y=disease.proportion, color=disease)) + geom_point(alpha=0.2, show.legend = FALSE) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data, show.legend =FALSE) + facet_grid(disease ~ .)

ggplot(na.omit(diseases), aes(x=year, y=disease.proportion, color=disease)) + geom_point(alpha=0.2, show.legend = FALSE) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data, show.legend =FALSE) + facet_wrap(~disease, scales = "free")

ggplot(na.omit(diseases), aes(x=year, y=normalized.count, color=disease)) + geom_boxplot(aes(group=factor(year)),show.legend = F, size=0.2, outlier.size = 0.05) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data, show.legend =FALSE) + facet_wrap(~disease, scales = "free")

ggplot(na.omit(diseases), aes(x=year, y=percentpop, color=disease)) + geom_boxplot(aes(group=factor(year)),show.legend = F, size=0.2, outlier.size = 0.05) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data, show.legend =FALSE) + facet_wrap(~disease, scales = "free")
```
```{r, fig.width=12, fig.height=4}
colors <- scales::hue_pal()(6)
colordf <- data.frame(disease=levels(diseases$disease), color=colors)
colorlist <- colordf[which(colordf$disease %in% c("Hepatitis A", "Measles", "Rubella")),"color"]
colorlist <- as.character(colorlist)

ggplot(na.omit(diseases[which(diseases$disease %in% c("Hepatitis A", "Measles", "Rubella")),]), aes(x=year, y=normalized.count, color=disease)) + geom_boxplot(aes(group=factor(year), fill=year), show.legend = TRUE, size=0.2, outlier.size = 0.1, outlier.alpha = 0.7) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Hepatitis A", "Measles", "Rubella")),], show.legend =FALSE, linetype=2, size=.3) + facet_wrap(~disease, scales = "free") + scale_color_manual(values = colorlist)
```




```{r, fig.width=12, fig.height=4}
require(gridExtra)
meas<- ggplot(na.omit(diseases[which(diseases$disease =="Measles"),]), aes(x=year, y=percentpop)) + geom_boxplot(aes(fill=factor(year)), color="#B79F00", show.legend = F, size=0.2, outlier.shape=NA, alpha=0.3) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Measles")),], show.legend =FALSE, linetype=2, size=.3, color="#B79F00") + facet_wrap(~disease, scales = "free") + coord_cartesian(ylim=c(0,0.0175))

hep<-ggplot(na.omit(diseases[which(diseases$disease =="Hepatitis A"),]), aes(x=year, y=percentpop)) + geom_boxplot(aes(fill=factor(year)), color="#00BFC4", show.legend = F, size=0.2, outlier.shape=NA, alpha=0.3) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Hepatitis A")),], color="#00BFC4", show.legend =FALSE, linetype=2, size=.3) + facet_wrap(~disease, scales = "free")+ coord_cartesian(ylim=c(0,0.0007))

rub<-ggplot(na.omit(diseases[which(diseases$disease =="Rubella"),]), aes(x=year, y=percentpop)) + geom_boxplot(aes(fill=factor(year)), color="#619CFF", show.legend = F, size=0.2, outlier.shape=NA, alpha=0.3) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Rubella")),], color="#619CFF", show.legend =FALSE, linetype=2, size=.3) + facet_wrap(~disease, scales = "free") + coord_cartesian(ylim=c(0,0.0013))
grid.arrange(meas,hep,rub, nrow=1)
```

Fill with mean for that year
```{r, fig.width=12, fig.height=4}
meandiseases <- diseases
newmean <- meandiseases %>%
  group_by(disease,year)%>%
  mutate(mean=round(mean(percentpop, na.rm = TRUE), digits=4))
newmean <- as.data.frame(newmean)

ggplot(na.omit(newmean[which(newmean$disease %in% c("Hepatitis A", "Measles", "Rubella")),]), aes(x=year, y=normalized.count, color=disease)) + geom_boxplot(aes(group=factor(year), fill=mean), show.legend = TRUE, size=0.2, outlier.size = 0.1, outlier.alpha = 0.7) + scale_colour_gradient() + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Hepatitis A", "Measles", "Rubella")),], show.legend =FALSE, linetype=2, size=.3) + facet_wrap(~disease, scales = "free") + scale_color_manual(values = colorlist) 
```

```{r, fig.width=12, fig.height=4}
require(gridExtra)
meas<- ggplot(na.omit(newmean[which(newmean$disease =="Measles"),]), aes(x=year, y=percentpop)) + geom_boxplot(aes(group=factor(year), fill=mean), color="darkgray", show.legend = F, size=0.1, outlier.shape=NA) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Measles")),], show.legend =FALSE, linetype=2, size=.3, color="#B79F00") + facet_wrap(~disease, scales = "free") + coord_cartesian(ylim=c(0,0.0175)) + theme(legend.position = "bottom")

hep<-ggplot(na.omit(newmean[which(newmean$disease =="Hepatitis A"),]), aes(x=year, y=percentpop)) + geom_boxplot(aes(group=factor(year), fill=mean), color="darkgray", show.legend = F, size=0.1, outlier.shape=NA) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Hepatitis A")),], color="#00BFC4", show.legend =FALSE, linetype=2, size=.3) + facet_wrap(~disease, scales = "free")+ coord_cartesian(ylim=c(0,0.0007))

rub<-ggplot(na.omit(newmean[which(newmean$disease =="Rubella"),]), aes(x=year, y=percentpop)) + geom_boxplot(aes(group=factor(year), fill=mean), color="darkgray", show.legend = F, size=0.1, outlier.shape=NA) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Rubella")),], color="#619CFF", show.legend =FALSE, linetype=2, size=.3) + facet_wrap(~disease, scales = "free") + coord_cartesian(ylim=c(0,0.0013))
grid.arrange(meas,hep,rub, nrow=1)
```


```{r, fig.width=10, fig.height=4}
levels <- c(seq(1919,2010,10), Inf)
labels <- seq(1920,2010,10)
factorbydecades <- diseases %>%
  mutate(decade = cut(year, levels, labels = labels))
  
ggplot(na.omit(factorbydecades[which(factorbydecades$disease %in% c("Hepatitis A", "Measles", "Rubella")),]), aes(x=decade, y=normalized.count, color=disease)) + geom_violin(aes(fill=factor(decade)), show.legend = F, alpha=0.3) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Hepatitis A", "Measles", "Rubella")),], show.legend =FALSE, linetype=2, size=.3) + facet_wrap(~disease, scales = "free") + scale_color_manual(values = colorlist)

ggplot(na.omit(factorbydecades[which(factorbydecades$disease %in% c("Hepatitis A", "Measles", "Rubella")),]), aes(x=decade, y=normalized.count, color=disease)) + geom_boxplot(aes(fill=factor(decade)), show.legend = F, alpha=0.3, size=0.3, outlier.size=.4) + facet_wrap(~ disease, scales="free") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Hepatitis A", "Measles", "Rubella")),], show.legend =FALSE, linetype=2, size=.3) + facet_wrap(~disease, scales = "free") + scale_color_manual(values = colorlist)
```

```{r}
ggplot(na.omit(diseases), aes(x=year, y=disease.proportion, color=disease)) + 
  geom_point(alpha=0.2) + 
  geom_vline(xintercept = hepA.vaccine, linetype="dashed", color = colors[1]) + 
  geom_vline(xintercept = measles.vaccine, linetype="dashed", color = colors[2]) +
  geom_vline(xintercept = mumps.vaccine, linetype="dashed", color = colors[3]) +
  geom_vline(xintercept = pertussis.vaccine, linetype="dashed", color = colors[4]) +
  geom_vline(xintercept = rubella.vaccine, linetype="dashed", color = colors[5]) 
  #+  geom_vline(xintercept = smallpox.vaccine, linetype="dashed", color = colors[6])
```

```{r}
ggplot(na.omit(diseases), aes(x=year, y=disease.proportion, color=disease)) + 
  geom_boxplot(aes(group=interaction(disease,factor(year))), show.legend = F, size=0.2, outlier.size = 0.1) +
  geom_vline(xintercept = hepA.vaccine, linetype="dashed", color = colors[1]) + 
  geom_vline(xintercept = measles.vaccine, linetype="dashed", color = colors[2]) +
  geom_vline(xintercept = mumps.vaccine, linetype="dashed", color = colors[3]) +
  geom_vline(xintercept = pertussis.vaccine, linetype="dashed", color = colors[4]) +
  geom_vline(xintercept = rubella.vaccine, linetype="dashed", color = colors[5]) 
  #+  geom_vline(xintercept = smallpox.vaccine, linetype="dashed", color = colors[6])
```


In 1930, Arizona, Utah, North Dakoata and Minnesota prohibited compulsory vaccination. Whereas, 10 states infant vaccination was compulsory for public schools including washington DC and Massachussets.

```{r}
ggplot(na.omit(diseases[which(diseases$state %in% c("Arizona", "Utah", "North Dakota", "Minnesota")),]), aes(x=year, y=disease.proportion, color=disease)) + 
  geom_point(alpha=0.2) + 
  geom_vline(xintercept = hepA.vaccine, linetype="dashed", color = colors[1]) + 
  geom_vline(xintercept = measles.vaccine, linetype="dashed", color = colors[2]) +
  geom_vline(xintercept = mumps.vaccine, linetype="dashed", color = colors[3]) +
  geom_vline(xintercept = pertussis.vaccine, linetype="dashed", color = colors[4]) +
  geom_vline(xintercept = polio.vaccine, linetype="dashed", color = colors[5]) +
  geom_vline(xintercept = rubella.vaccine, linetype="dashed", color = colors[6]) 

ggplot(na.omit(diseases[which(diseases$state %in% c("Massachusetts")),]), aes(x=year, y=disease.proportion, color=disease)) + 
  geom_point(alpha=0.2) + 
  geom_vline(xintercept = hepA.vaccine, linetype="dashed", color = colors[1]) + 
  geom_vline(xintercept = measles.vaccine, linetype="dashed", color = colors[2]) +
  geom_vline(xintercept = mumps.vaccine, linetype="dashed", color = colors[3]) +
  geom_vline(xintercept = pertussis.vaccine, linetype="dashed", color = colors[4]) +
  geom_vline(xintercept = polio.vaccine, linetype="dashed", color = colors[5]) +
  geom_vline(xintercept = rubella.vaccine, linetype="dashed", color = colors[6]) 
```


Hmm what happens if we group all the states
```{r}
us.diseases <- diseases %>%
  group_by(disease,year) %>%
  summarize(normalized.counts=sum(normalized.count, na.rm=T))
us.diseases
```
```{r}
ggplot(us.diseases, aes(x=year, y=normalized.counts, color=disease)) + 
  geom_point(alpha=0.5) + 
  geom_vline(xintercept = hepA.vaccine, linetype="dashed", color = colors[1]) + 
  geom_vline(xintercept = measles.vaccine, linetype="dashed", color = colors[2]) +
  geom_vline(xintercept = mumps.vaccine, linetype="dashed", color = colors[3]) +
  geom_vline(xintercept = pertussis.vaccine, linetype="dashed", color = colors[4]) +
  geom_vline(xintercept = polio.vaccine, linetype="dashed", color = colors[5]) +
  geom_vline(xintercept = rubella.vaccine, linetype="dashed", color = colors[6]) 
```

What about converting these to a proportion to emphasize vaccine effect (similar to plotting separately)
```{r}
maximums2 <- aggregate(normalized.counts~disease, us.diseases, FUN=max)
us.diseases$maximums <- maximums2[us.diseases$disease, "normalized.counts"]
us.diseases$disease.proportion <- us.diseases$normalized.counts/us.diseases$maximums
```

```{r}
ggplot(us.diseases, aes(x=year, y=disease.proportion, color=disease)) + 
  geom_point(alpha=0.5) + 
  geom_vline(xintercept = hepA.vaccine, linetype="dashed", color = colors[1]) + 
  geom_vline(xintercept = measles.vaccine, linetype="dashed", color = colors[2]) +
  geom_vline(xintercept = mumps.vaccine, linetype="dashed", color = colors[3]) +
  geom_vline(xintercept = pertussis.vaccine, linetype="dashed", color = colors[4]) +
  geom_vline(xintercept = polio.vaccine, linetype="dashed", color = colors[5]) +
  geom_vline(xintercept = rubella.vaccine, linetype="dashed", color = colors[6]) 
```

Some of these vaccines were created before we have data for. Let's look at individual plots of those diseases that we have data before the vaccine was created

I want to try to fit a piecewise linear model on Hepatitis A.

```{r}
HepAPW <- lm(normalized.count ~ year, data=diseases[which(diseases$disease == "Hepatitis A"),])
segmented.mod <- segmented(HepAPW, seg.Z = ~year, psi=hepA.vaccine)
plot(diseases[which(diseases$disease == "Hepatitis A"),"year"], diseases[which(diseases$disease == "Hepatitis A"),"normalized.count"])
plot(segmented.mod, add=T)
```

that estimates the break way to early
```{r}
library(splines)
hepplot <- ggplot(na.omit(diseases[which(diseases$disease == "Hepatitis A"),]), aes(x=year, y=normalized.count, color=disease)) + geom_point(alpha=0.2, show.legend = FALSE) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease == "Hepatitis A"),], show.legend =FALSE)

lm1 <- lm(normalized.count ~ bs(year, df = NULL, knots  = hepA.vaccine, degree=1), data=diseases[which(diseases$disease == "Hepatitis A"),])
# create predictions with this lm
newdat <- data.frame(year = seq(from = min(diseases[which(diseases$disease == "Hepatitis A"),"year"]), to = max(diseases[which(diseases$disease == "Hepatitis A"),"year"]), by = 1))
newdat$normalized.count <- predict(lm1, newdata = newdat)
newdat$disease <- "Hepatitis A"

# add to the plot
hepplot + geom_line(data=newdat, aes(year, normalized.count), color="black", linetype=3)
```

```{r}
diseaseofinterest <- "Measles"
colorofinterest <- colors[2]
vaccine <- measles.vaccine
newplot <- ggplot(na.omit(diseases[which(diseases$disease == diseaseofinterest),]), aes(x=year, y=normalized.count)) + geom_point(alpha=0.2, show.legend = FALSE, color=colorofinterest) + geom_vline(aes(xintercept=vl), data=vline.data[which(vline.data$disease == diseaseofinterest),], show.legend =FALSE, color=colorofinterest)

lm1 <- lm(normalized.count ~ bs(year, df = NULL, knots  = vaccine, degree=1), data=diseases[which(diseases$disease == diseaseofinterest),])
# create predictions with this lm
newdat <- data.frame(year = seq(from = min(diseases[which(diseases$disease == diseaseofinterest),"year"]), to = max(diseases[which(diseases$disease == diseaseofinterest),"year"]), by = 1))
newdat$normalized.count <- predict(lm1, newdata = newdat)
newdat$disease <- diseaseofinterest

# add to the plot
newplot + geom_line(data=newdat, aes(year, normalized.count), color="black", linetype=3)
```

Show it on all of them
```{r}
piecewisevals <- function(vax, doi){
  newlm <- lm(normalized.count ~ bs(year, df = NULL, knots  = vax, degree=1), data=diseases[which(diseases$disease == doi),])
  newdat <- data.frame(year = seq(from = min(diseases[which(diseases$disease == doi),"year"]), to = max(diseases[which(diseases$disease == doi),"year"]), by = 1))
  newdat$normalized.count <- predict(newlm, newdata = newdat)
  newdat$disease <- doi
  newdat
}

piecewise <- data.frame(disease=character(),
                        year=numeric(),
                        normalized.count=numeric())
for (dis in levels(diseases$disease)){
  vdate <- vline.data[which(vline.data$disease == dis),"vl"]
  dftoadd <- piecewisevals(vdate, dis)
  piecewise <- rbind(piecewise, dftoadd)
}
```

```{r}
facetedplot <- ggplot(na.omit(diseases), aes(x=year, y=normalized.count, color=disease)) + geom_point(alpha=0.2, show.legend = FALSE) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease != "Smallpox"),], show.legend =FALSE) + facet_grid(disease~., scales="free")
facetedplot + geom_line(data=piecewise, aes(year, normalized.count), color = "black", linetype=3)
```
Let's look at disease progression 10 years before the vaccine and 10 years after
```{r}
piecewisevals2 <- function(vax, doi){
  newlm <- lm(normalized.count ~ bs(year, df = NULL, knots  = vax, degree=1), data=diseases[which(diseases$disease == doi & diseases$year > vax-10 & diseases$year <vax+10),])
  newdat <- data.frame(year = seq(from = vax-10, to = vax+10, by = 1))
  newdat
  newdat$normalized.count <- predict(newlm, newdata = newdat)
  newdat$disease <- doi
  newdat
}

piecewise2 <- data.frame(disease=character(),
                        year=numeric(),
                        normalized.count=numeric())
for (dis in unique(diseases[which(diseases$disease %in% list("Hepatitis A", "Measles", "Rubella")),"disease"])){
  vdate <- vline.data[which(vline.data$disease == dis),"vl"]
  dftoadd <- piecewisevals2(vdate, dis)
  piecewise2 <- rbind(piecewise2, dftoadd)
}
```

```{r}
facetedplot <- ggplot(na.omit(diseases[which(diseases$disease %in% list("Hepatitis A", "Measles", "Rubella")),]), aes(x=year, y=normalized.count, color=disease)) + geom_point(alpha=0.2, show.legend = FALSE) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease != "Smallpox" & vline.data$disease != "Pertussis" & vline.data$disease != "Mumps"),], show.legend =FALSE) + facet_grid(disease~., scales="free")
facetedplot + geom_line(data=piecewise2, aes(year, normalized.count), color = "black", linetype=3)
```

Let's subset the data for the 10 years before and after:
```{r}
diseasesofinterest = c("Hepatitis A", "Measles", "Rubella")
tmpdf <- diseases[0,]
for (disease in diseasesofinterest){
  vaccineyear <- vline.data[which(vline.data$disease == disease),"vl"]
  tmpdf2 <- diseases[which(diseases$disease == disease & diseases$year > vaccineyear-10 & diseases$year < vaccineyear+10),]
  tmpdf <- rbind(tmpdf, tmpdf2)
}
tmpdf$disease <- factor(tmpdf$disease)
vaccinediseases <- tmpdf
vaccinediseases$vaccineyear <- vline.data$vl[match(vaccinediseases$disease, vline.data$disease)]
vaccinediseases$yearssincevaccine <- vaccinediseases$year - vaccinediseases$vaccineyear
```

check it out
```{r}
ggplot(na.omit(vaccinediseases), aes(x=yearssincevaccine, y=normalized.count, color=disease)) + geom_point(alpha=0.2, show.legend = FALSE) + geom_vline(aes(xintercept=0, color=disease), data=vline.data[which(vline.data$disease %in% diseasesofinterest),], show.legend =FALSE) + facet_grid(disease~., scales="free")
```

Okay add the piecewise
```{r}
piecewisevals3 <- function(vax, doi){
  newlm <- lm(normalized.count ~ bs(year, df = NULL, knots  = vax, degree=1), data=vaccinediseases[which(vaccinediseases$disease == doi),])
  newdat <- data.frame(year = seq(from = min(vaccinediseases[which(vaccinediseases$disease == doi),"year"]), to = max(vaccinediseases[which(vaccinediseases$disease == doi),"year"]), by = 1))
  newdat$normalized.count <- predict(newlm, newdata = newdat)
  newdat$disease <- doi
  newdat
}

piecewise3 <- data.frame(disease=character(),
                        year=numeric(),
                        normalized.count=numeric())
for (dis in levels(vaccinediseases$disease)){
  vdate <- vline.data[which(vline.data$disease == dis),"vl"]
  dftoadd <- piecewisevals3(vdate, dis)
  piecewise3 <- rbind(piecewise3, dftoadd)
}
```

```{r, fig.width=8, fig.height=4}
ggplot(na.omit(vaccinediseases), aes(x=year, y=normalized.count, color=disease)) + geom_point(alpha=0.2, show.legend = FALSE) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% diseasesofinterest),], show.legend =FALSE, linetype=3) + facet_wrap(~disease, scales="free") + geom_line(data=piecewise3, aes(year, normalized.count), color = "black")
```

How does boxplots look?
```{r, fig.width=12, fig.height=4}
ggplot(na.omit(vaccinediseases), aes(x=year, y=normalized.count, color=disease)) + geom_violin(aes(fill=factor(year)),size=.2,show.legend = FALSE, color="darkgray") + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% diseasesofinterest),], show.legend =FALSE, linetype="dotdash") + facet_wrap(~disease, scales="free") + geom_line(data=piecewise3, aes(year, normalized.count, color=disease), show.legend = F)+ scale_color_manual(values = colorlist)
```

to color by the mean
```{r}
vaccinediseases <- vaccinediseases %>%
  group_by(disease,year)%>%
  mutate(mean=round(mean(normalized.count, na.rm = TRUE)))
vaccinediseases <- as.data.frame(vaccinediseases)
```

```{r, fig.width=12, fig.height=4}
require(gridExtra)
meas<- ggplot(na.omit(vaccinediseases[which(vaccinediseases$disease =="Measles"),]), aes(x=year, y=normalized.count)) + geom_violin(aes(group=factor(year), fill=mean), trim=TRUE, color="#B79F00", show.legend = F, size=.3) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Measles")),], show.legend =FALSE, linetype=2, size=.3, color="black") + geom_line(data=piecewise3[which(piecewise3$disease == "Measles"),], aes(year, normalized.count), color="#B79F00", show.legend = F) + facet_wrap(~disease, scales = "free") + coord_cartesian(ylim=c(0,15000)) + theme(legend.position = "bottom")

hep<-ggplot(na.omit(vaccinediseases[which(vaccinediseases$disease =="Hepatitis A"),]), aes(x=year, y=normalized.count)) + geom_violin(aes(group=factor(year), fill=mean), trim=TRUE, color="#00BFC4", show.legend = F, size=.3) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Hepatitis A")),], color="black", show.legend =FALSE, linetype=2, size=.3) + geom_line(data=piecewise3[which(piecewise3$disease == "Hepatitis A"),], aes(year, normalized.count), color="#00BFC4", show.legend = F) + facet_wrap(~disease, scales = "free") + facet_wrap(~disease, scales = "free")+ coord_cartesian(ylim=c(0,600))+ ylab(NULL)

rub<-ggplot(na.omit(vaccinediseases[which(vaccinediseases$disease =="Rubella"),]), aes(x=year, y=normalized.count)) + geom_violin(aes(group=factor(year), fill=mean, trim=TRUE), color="#619CFF", show.legend = F, size=.3) + geom_vline(aes(xintercept=vl, color=disease), data=vline.data[which(vline.data$disease %in% c("Rubella")),], color="black", show.legend =FALSE, linetype=2, size=.3) + geom_line(data=piecewise3[which(piecewise3$disease == "Rubella"),], aes(year, normalized.count), color="#619CFF", show.legend = F) + facet_wrap(~disease, scales = "free") + facet_wrap(~disease, scales = "free") + coord_cartesian(ylim=c(0,1250)) + ylab(NULL)
grid.arrange(meas,hep,rub, nrow=1)
```
